\chapter{Logging and debugging strategies}
\label{ch:logging}

\begin{aside}
A TUI cannot print to stderr without messing up the screen.
But you still need a way to see what's happening when
something goes wrong.
\end{aside}

\section{Log to a file}

\maya{}'s default logger writes to \code{maya.log} in the
current directory:

\begin{lstlisting}
LOG_INFO("Started session");
LOG_ERROR("Failed to load: {}", err);
\end{lstlisting}

Format strings via \code{fmt::format} or
\code{std::format}.

\section{Levels}

ERROR, WARN, INFO, DEBUG, TRACE. Filter via environment:

\begin{lstlisting}
MAYA_LOG=DEBUG ./myapp
\end{lstlisting}

\section{Tail in another terminal}

\begin{lstlisting}
tail -f maya.log
\end{lstlisting}

The log scrolls as your app runs. This is the standard
debug workflow.

\section{Structured logging}

For machine consumption, JSON Lines:

\begin{lstlisting}
{"ts":"2024-01-01T12:00:00Z","level":"INFO","msg":"started"}
\end{lstlisting}

Parsable by jq, log aggregators, etc.

\section{In-app log view}

For users to see logs without leaving the app, build a
log-tail widget:

\begin{lstlisting}
auto log = file_tail("maya.log") | scrollable;
\end{lstlisting}

Bound to a hotkey like F10.

\section{Conditional breakpoints}

Inside a debugger, set a condition on a log line:

\begin{lstlisting}
break log.cpp:42 if level == ERROR
\end{lstlisting}

Catches errors only.

\section{Crash dumps}

Install a signal handler that writes a stack trace to a
file before re-raising. \maya{} ships a default that uses
\code{libbacktrace}.

\section{Reproducing bugs}

Save a session trace (every input event) and replay:

\begin{lstlisting}
MAYA_TRACE=session.trace ./myapp        # record
./myapp --replay session.trace          # replay
\end{lstlisting}

\section{Recap}

\begin{recap}
\begin{itemize}[leftmargin=*]
  \item Log to file, not stderr.
  \item Levels filterable via env.
  \item Tail in a side terminal.
  \item Structured logs for machine reading.
  \item Crash dumps for after-the-fact analysis.
  \item Record + replay for bug reproduction.
\end{itemize}
\end{recap}

\section*{Workshop}

\begin{enumerate}[leftmargin=*]
  \item Install the logger; tail in another terminal.
  \item Build an in-app log viewer bound to F10.
  \item Wire a crash dump handler.
\end{enumerate}
