\chapter{Networking and remote state}
\label{ch:networking}

\begin{aside}
Many TUIs are clients: a chat app, an RSS reader, an SSH
multiplexer. Networking adds latency, errors, and a need to
think about offline.
\end{aside}

\section{Sockets}

\maya{} doesn't bundle a network library --- use what you
need. \code{asio}, \code{boost.asio}, \code{libuv} for
async I/O; raw sockets for simple HTTP.

\section{Integration with the loop}

The event loop's \code{poll} can watch multiple fds. Add
the socket to the watch list; on readable, drain into a
buffer; parse into events.

\begin{lstlisting}
ctx.watch_fd(sock, [&] {
    auto bytes = read_some(sock);
    parse_protocol(bytes);
});
\end{lstlisting}

\section{Offline UI}

Network requests fail. The UI should:

\begin{itemize}[leftmargin=*]
  \item Show a connection state indicator.
  \item Queue actions while offline; retry on reconnect.
  \item Display cached data if no fresh data.
  \item Never freeze.
\end{itemize}

\section{Retry and backoff}

Exponential backoff: 1 s, 2 s, 4 s, 8 s, capped. Add
jitter to avoid thundering herd.

\section{HTTP basics}

\maya{} apps doing simple HTTP can shell out to \code{curl}
or use a small library (\code{cpr}, \code{httplib}).

\section{WebSockets}

For server-pushed updates, WebSockets. \code{IXWebSocket},
\code{websocketpp} are options.

\section{State sync}

Two strategies:

\begin{itemize}[leftmargin=*]
  \item \textbf{Pull.} Poll periodically.
  \item \textbf{Push.} Server pushes updates over a
    persistent connection.
\end{itemize}

Push is gentler on the wire and more responsive; harder to
implement.

\section{Optimistic UI}

Update local state immediately on user action; reconcile
with server response. If the server disagrees, roll back
visually.

\section{Recap}

\begin{recap}
\begin{itemize}[leftmargin=*]
  \item Watch sockets via the same poll loop.
  \item Offline indicator + retry queue + cached data.
  \item Exponential backoff with jitter.
  \item Push beats pull for responsiveness.
  \item Optimistic UI feels fast.
\end{itemize}
\end{recap}

\section*{Workshop}

\begin{enumerate}[leftmargin=*]
  \item Wire a socket into the event loop.
  \item Build a connection-state widget.
  \item Implement optimistic update with rollback.
\end{enumerate}
