\chapter{Worker pool and thread bridge}
\label{ch:workers}

\begin{aside}
The main thread paints. Workers compute. The bridge is a
queue: tasks go in, results come back to the main thread for
signal updates.
\end{aside}

\section{The pool}

\maya{} ships a small thread pool sized to
\code{std::thread::hardware\_concurrency()}. Each worker
pulls tasks from a lock-free queue.

\begin{lstlisting}
ThreadPool pool(N);
pool.submit([] { heavy_work(); });
\end{lstlisting}

\section{Main-thread post-back}

A worker that wants to update a signal posts back:

\begin{lstlisting}
pool.submit([&] {
    auto result = compute();
    ctx.post_to_main([result] {
        ctx.set(value, result);
    });
});
\end{lstlisting}

\code{post\_to\_main} pushes a microtask onto the main
loop's queue; the loop drains it before the next render.

\section{Why not direct signal writes}

The signal graph is single-threaded. A worker writing a
signal would race with the main thread reading it. Posting
back serialises the update.

\section{Queues}

The main-thread queue is bounded (e.g., 1024 entries). A
worker that fills it backs off briefly. This is rare in
practice --- the main thread drains fast.

\section{Backpressure}

Workers should not produce results faster than the main
thread can consume. For high-throughput jobs (image
decoding bursts), batch results:

\begin{lstlisting}
pool.submit([&] {
    std::vector<Result> batch;
    for (int i = 0; i < 100; ++i)
        batch.push_back(work(i));
    post_to_main([batch] {
        for (auto& r : batch) ctx.set(...);
    });
});
\end{lstlisting}

One queue entry, many results.

\section{Long-running workers}

For continuous tasks (file watcher, network listener), the
worker loops on its own and posts results as they arrive.
The pool's submit is for one-shot tasks; long-runners use a
dedicated thread.

\section{Cancellation}

Worker tasks check a cancel flag periodically and bail
early. The cancel flag is owned by the application; the
worker just reads it.

\section{Recap}

\begin{recap}
\begin{itemize}[leftmargin=*]
  \item Pool sized to hardware concurrency.
  \item Workers post back to the main thread for signal
    updates.
  \item Bounded queues with backpressure.
  \item Batch results when high-rate.
  \item Long-running tasks need dedicated threads.
\end{itemize}
\end{recap}

\section*{Workshop}

\begin{enumerate}[leftmargin=*]
  \item Build a parallel image decoder. Submit one task
    per image; observe parallelism.
  \item Add backpressure by counting in-flight tasks;
    refuse to submit if too many.
  \item Build a file-watcher worker that posts changes
    back to the main thread.
\end{enumerate}
