\chapter{Internationalisation}
\label{ch:i18n2}

\begin{aside}
Hard-coded English makes your app readable only to English
speakers. Internationalisation lets the same code serve a
French user, a Japanese user, a right-to-left Arabic user.
\end{aside}

\section{Message catalogues}

Strings live in catalogues, one per locale:

\begin{lstlisting}
// en.toml
greeting = "Hello, {}"

// fr.toml
greeting = "Bonjour, {}"
\end{lstlisting}

\section{Lookup}

\begin{lstlisting}
auto msg = i18n("greeting", username);
\end{lstlisting}

Lookup picks the active locale's catalogue, substitutes
arguments.

\section{Pluralisation}

``1 file'' vs ``2 files''. CLDR plural rules:

\begin{lstlisting}
auto msg = i18n_plural("file_count", n);
// en: "1 file" / "{} files"
// ru: 3 forms based on n
\end{lstlisting}

\code{libicu} or smaller libs handle the rules.

\section{Date and time formatting}

Locale-dependent: ``Jan 1, 2024'' vs ``1 janvier 2024''.
Use \code{std::format} with locale, or icu, or a small
manual table.

\section{Number formatting}

Thousands separator: \code{,} vs \code{.} vs space.
Decimal: \code{.} vs \code{,}. Use locale-aware formatters.

\section{Right-to-left}

Arabic, Hebrew flow right-to-left. \maya{} doesn't shape
RTL automatically (see Chapter~\ref{ch:shaping}). For
RTL-only sections, pre-reverse and use the result; for
mixed, you need a BiDi library.

\section{Font and width}

Some scripts (Devanagari, Thai) need cluster shaping that
the terminal may not do well. Test on representative
fonts.

\section{Locale detection}

Check \code{LANG} and \code{LC\_ALL} environment
variables. Fallback to en\_US.UTF-8.

\section{Recap}

\begin{recap}
\begin{itemize}[leftmargin=*]
  \item Catalogues per locale.
  \item Pluralisation via CLDR.
  \item Date, time, number locale-aware.
  \item RTL needs care.
  \item Detect via \code{LANG}; default to English.
\end{itemize}
\end{recap}

\section*{Workshop}

\begin{enumerate}[leftmargin=*]
  \item Add an English and French catalogue; switch via
    config.
  \item Implement plural messages.
  \item Test your app with \code{LANG=fr\_FR.UTF-8}.
\end{enumerate}
