\chapter{Profiling and optimisation}
\label{ch:profile}

\begin{aside}
Before optimising, measure. After measuring, fix the worst.
After fixing, measure again. The cycle ends when the slowest
part is fast enough.
\end{aside}

\section{Tools}

\begin{itemize}[leftmargin=*]
  \item \code{perf}: Linux sampling profiler.
  \item \code{Tracy}: instrumentation profiler.
  \item \code{Instruments}: macOS.
  \item \code{callgrind}: cache misses, branch
    mispredicts.
  \item \code{vtune}: Intel, deep micro-arch.
\end{itemize}

\section{Sampling vs instrumentation}

Sampling (\code{perf}): periodic stack traces. Low
overhead, identifies hot functions.

Instrumentation (\code{Tracy}): trace points in code, full
detail. Higher overhead, gives complete picture.

For exploration, start with sampling. For deep tuning,
switch to instrumentation.

\section{What to measure}

In \maya{}, the typical hot spots are:

\begin{itemize}[leftmargin=*]
  \item Diff inner loop (SIMD or scalar).
  \item Element-tree walk (layout).
  \item ANSI emission.
  \item Signal-graph propagation.
\end{itemize}

If your hot spot is application code (e.g., string
formatting), focus there.

\section{Common wins}

\begin{itemize}[leftmargin=*]
  \item Inline hot functions.
  \item Reduce allocations (arena, reuse).
  \item Cache derived values (memos).
  \item Vectorise inner loops.
  \item Reduce branch mispredicts.
\end{itemize}

\section{Compiler flags}

\begin{itemize}[leftmargin=*]
  \item \code{-O3}: aggressive optimisation.
  \item \code{-march=native}: target the build machine's
    ISA.
  \item \code{-flto}: link-time optimisation.
  \item \code{-fno-omit-frame-pointer}: for profilers.
\end{itemize}

\section{PGO}

Profile-guided optimisation: build instrumented, run a
representative workload, rebuild with the profile data.
Gains of 10--20\% are typical.

\section{When to stop}

A profile dominated by ``main loop'' and ``poll'' means the
work itself is small. Time goes to waiting; the CPU is
already efficient.

\section{Recap}

\begin{recap}
\begin{itemize}[leftmargin=*]
  \item Sampling for exploration; instrumentation for
    detail.
  \item Hot spots: diff, layout, emit, signals.
  \item Common wins: inline, reduce alloc, cache, vectorise.
  \item PGO for the last 10--20\%.
\end{itemize}
\end{recap}

\section*{Workshop}

\begin{enumerate}[leftmargin=*]
  \item Profile your app with \code{perf}; identify the
    hottest function.
  \item Try \code{-flto} and \code{-march=native}; measure
    the gain.
  \item Set up PGO for one representative workload.
\end{enumerate}
