\chapter{Memory model and ownership}
\label{ch:memory}

\begin{aside}
\maya{} avoids garbage collection and reference counting. The
ownership tree is the memory model. Each owner deallocates
its tree on destruction.
\end{aside}

\section{Ownership tree}

Every signal, effect, store, and widget has an owner. Owners
form a tree rooted at the application context. When an owner
dies, everything it owns dies with it.

This mirrors lexical scoping: variables live as long as the
scope they were declared in.

\section{Smart pointers}

\maya{} uses:

\begin{itemize}[leftmargin=*]
  \item \code{unique\_ptr} for sole ownership.
  \item \code{shared\_ptr} sparingly --- only when ownership
    is genuinely shared and the cycle risk is bounded.
  \item Raw pointers for non-owning views.
\end{itemize}

\section{Arena allocation}

The element tree lives in an arena (Chapter~\ref{ch:allocations}).
Reset between frames. No individual destructors run --- the
arena resets and forgets. So arena-allocated types must be
trivially destructible.

\section{Strict aliasing}

\maya{} doesn't violate strict aliasing. Cells are
\code{uint64\_t}; cell-meaning is achieved by bit unpacking,
not pointer reinterpretation.

\section{Concurrency model}

Single-threaded UI thread by default. Workers post results
back. Signals are not thread-safe by design.

\code{TSan} run on the test suite verifies no races.

\section{Move semantics}

DSL builders use rvalue references. Element construction
is move-only; no copying of element trees.

\section{Lifetime annotations}

\cpp{}26 lifetime annotations are not yet broadly available
in compilers. \maya{} uses comments and code review for
ownership clarity until they are.

\section{Recap}

\begin{recap}
\begin{itemize}[leftmargin=*]
  \item Ownership tree, not GC or refcount.
  \item Arena for element-tree memory.
  \item Single-threaded UI; workers post back.
  \item Move semantics for cheap composition.
\end{itemize}
\end{recap}

\section*{Workshop}

\begin{enumerate}[leftmargin=*]
  \item Trace your app's ownership tree. Identify any
    smart pointer that could be replaced with raw.
  \item Run under TSan; fix any race.
  \item Audit the element tree for non-trivial
    destructors that snuck into arena allocation.
\end{enumerate}
