\chapter{Testing strategies for TUIs}
\label{ch:testing2}

\begin{aside}
A TUI is hard to test because the output is bytes flowing
through a pty. Three layers help: unit, snapshot, integration.
Each catches a different class of bug.
\end{aside}

\section{Unit tests}

Pure logic: signal updates, reducers, formatters. No
rendering. Standard \cpp{} unit-test frameworks (Catch2,
doctest, GoogleTest).

\begin{lstlisting}
TEST_CASE("counter increments") {
    auto c = signal(0);
    increment(c);
    REQUIRE(get(c) == 1);
}
\end{lstlisting}

\section{Snapshot tests}

See Chapter~\ref{ch:snapshot}. Render to a string backend,
compare against a golden.

\section{Integration tests}

Drive the full app with simulated input:

\begin{lstlisting}
TEST_CASE("login flow") {
    auto app = TestApp();
    app.type("alice");
    app.tab();
    app.type("secret");
    app.enter();
    REQUIRE(app.is_on("/home"));
}
\end{lstlisting}

The harness sends keys, advances the clock, awaits idle,
then asserts on state.

\section{Property tests}

Random input streams to verify invariants:

\begin{lstlisting}
PROPERTY("never crashes on random keys") {
    auto app = TestApp();
    auto keys = arbitrary_key_sequence();
    for (auto k : keys) app.send(k);
    REQUIRE(!app.crashed());
}
\end{lstlisting}

Rapidcheck, hypothesis. Find edge cases humans miss.

\section{Fuzzing}

libFuzzer or AFL drive your input parser with random
bytes. Catches buffer overruns and parser bugs.

\section{Visual regression}

CI captures screenshots; compare pixel-by-pixel. Only
useful when running in a fixed terminal emulator.

\section{Coverage}

\code{gcov} or \code{lcov} measure which lines run.
Target 80\%+ coverage of business logic; lower for
boilerplate is acceptable.

\section{Recap}

\begin{recap}
\begin{itemize}[leftmargin=*]
  \item Unit for logic, snapshot for output, integration
    for flows.
  \item Property tests for randomness.
  \item Fuzz the input parser.
  \item Coverage as a guide, not a goal.
\end{itemize}
\end{recap}

\section*{Workshop}

\begin{enumerate}[leftmargin=*]
  \item Write a unit test, a snapshot test, and an
    integration test for one feature.
  \item Add a property test that exercises random keys.
  \item Measure coverage; identify untested paths.
\end{enumerate}
