\chapter{Custom widgets from scratch}
\label{ch:customwidget}

\begin{aside}
The built-in widgets cover 80\% of cases. The other 20\% is
you. Custom widgets are a small protocol: measure, layout,
paint, handle events.
\end{aside}

\section{The widget protocol}

\begin{lstlisting}
class MyWidget : public Widget {
public:
    Size measure(Constraints c) override;
    void layout(Rect bounds) override;
    void paint(Painter& p) override;
    bool on_event(Event e) override;
};
\end{lstlisting}

Four methods. Each runs at a specific phase of the frame.

\section{Measure}

Given constraints (min/max width and height), report your
preferred size. Layout uses this to allocate.

\section{Layout}

You are told your final rectangle. Position any children
within it.

\section{Paint}

Write cells into the painter. Respect the clip rectangle
(automatic).

\section{Events}

Receive keyboard, mouse, focus events. Return true to
consume; false to let parent handle.

\section{State}

Per-instance state lives in member variables. Persistent
state across frames (via the widget identity in
Chapter~\ref{ch:reconcile}) survives rebuilds.

\section{A complete example: progress ring}

\begin{lstlisting}
class ProgressRing : public Widget {
    float pct;
public:
    Size measure(Constraints c) override {
        return {5, 3};
    }
    void layout(Rect b) override { bounds = b; }
    void paint(Painter& p) override {
        int fill = std::round(pct * 4);
        const char* glyphs[] =
            {" . o O *"};
        p.put(bounds.x + 2, bounds.y + 1,
              Cell{ glyphs[fill] });
    }
};
\end{lstlisting}

A 5x3 widget showing one of 5 fill levels.

\section{Composition}

A complex widget composes simpler ones rather than drawing
from scratch. Wrap a text element inside a border element
inside an alignment element.

Drop to raw paint only when no composition works.

\section{Recap}

\begin{recap}
\begin{itemize}[leftmargin=*]
  \item Widget = measure + layout + paint + event.
  \item Compose where possible; raw paint when not.
  \item State in members; identity from tree path.
\end{itemize}
\end{recap}

\section*{Workshop}

\begin{enumerate}[leftmargin=*]
  \item Build a custom widget from scratch.
  \item Add it to your app and use it.
  \item Refactor: can it be composed from built-ins
    instead?
\end{enumerate}
