\chapter{Buttons and clickables}
\label{ch:buttons}

\begin{aside}
A button is the simplest interactive widget: text plus an
action. The work is in the polish --- hover, press, disabled,
focus ring, keyboard activation.
\end{aside}

\section{The minimal button}

\begin{lstlisting}
auto btn = text("Save") | on_click(save_action);
\end{lstlisting}

A click anywhere on the text fires the action. That's it.

\section{States}

Buttons have visual states:

\begin{itemize}[leftmargin=*]
  \item \textbf{Normal.} Default appearance.
  \item \textbf{Hover.} Mouse over, slightly different.
  \item \textbf{Pressed.} Mouse down, inverted or
    darkened.
  \item \textbf{Focused.} Selected via keyboard, ring or
    different style.
  \item \textbf{Disabled.} Greyed out, not clickable.
\end{itemize}

\maya{}'s button widget tracks each via signals.

\section{Keyboard activation}

Focused button + Enter (or Space) fires the action. The
default key handler:

\begin{lstlisting}
on_key(KeyEnter, [&] { if (focused) action(); });
\end{lstlisting}

\section{Focus ring}

The focus ring is a subtle border or background change.
Honour OS conventions: dotted underline on Mac, solid
border on Windows, custom on Linux.

\maya{}'s default is a 1px bright border. Override per
theme.

\section{Toggle buttons}

A toggle button has on/off state:

\begin{lstlisting}
auto toggle = button("Mute")
            | toggleable(is_muted);
\end{lstlisting}

The signal tracks the state; the visual reflects it.

\section{Button groups}

A row of buttons where only one can be active (radio
group):

\begin{lstlisting}
auto group = radio_group(options, selected);
\end{lstlisting}

Internally each button is a toggle bound to the selected
signal.

\section{Confirmation buttons}

A dangerous action (delete, sign out) deserves a
confirmation:

\begin{lstlisting}
auto del = button("Delete")
         | confirm("Are you sure?");
\end{lstlisting}

The button opens a modal; the action fires only on the
modal's confirm.

\section{Async actions}

A button that triggers async work should show progress:

\begin{lstlisting}
auto btn = button(loading ? "Saving..." : "Save")
         | disabled(loading)
         | on_click([&]() -> Task<void> {
             ctx.set(loading, true);
             co_await save();
             ctx.set(loading, false);
           });
\end{lstlisting}

\section{Recap}

\begin{recap}
\begin{itemize}[leftmargin=*]
  \item Button = text + click action.
  \item Five visual states.
  \item Keyboard activation via Enter or Space.
  \item Toggle, radio, confirmation, async variants.
\end{itemize}
\end{recap}

\section*{Workshop}

\begin{enumerate}[leftmargin=*]
  \item Build a button with all five visual states.
  \item Wire keyboard activation for focused buttons.
  \item Add async loading state to a save button.
\end{enumerate}
