\chapter{Animation curves and physics}
\label{ch:curves}

\begin{aside}
A linear animation feels mechanical. A spring feels alive.
The math is simple; the difference is huge.
\end{aside}

\section{Curves library}

\maya{} ships:

\begin{itemize}[leftmargin=*]
  \item Linear.
  \item Quadratic ease-in / ease-out / ease-in-out.
  \item Cubic, quartic, quintic variants.
  \item Sine, exponential, circular.
  \item Bounce, elastic, back.
  \item Spring (physics-based).
\end{itemize}

\section{Cubic bezier}

The CSS bezier curve takes four control points:

\begin{lstlisting}
auto curve = bezier(0.25, 0.1, 0.25, 1.0);  // ease
\end{lstlisting}

Most named curves are bezier presets.

\section{Spring physics}

A spring is parameterised by stiffness, damping, mass:

\begin{lstlisting}
struct Spring {
    float stiffness = 170;
    float damping = 26;
    float mass = 1;
    float advance(float dt);
};
\end{lstlisting}

The animation runs until the spring settles (velocity
near zero). Duration is not fixed --- it depends on
parameters.

\section{When to use what}

\begin{itemize}[leftmargin=*]
  \item Linear: progress bars (literal motion).
  \item Ease-out: things appearing (UI feels polished).
  \item Ease-in: things disappearing.
  \item Spring: organic motion (drag-and-drop snap-back).
  \item Bounce: playful only; avoid for tools.
\end{itemize}

\section{Per-axis curves}

A 2-D animation can use different curves per axis. Slide
horizontally with ease-out, fade with linear.

\section{Performance}

Curve evaluation is one or two multiplies per frame. Even
hundreds of concurrent animations cost microseconds.

\section{Reduced motion}

When the user prefers reduced motion, snap to the end
value. The curve is bypassed entirely.

\section{Recap}

\begin{recap}
\begin{itemize}[leftmargin=*]
  \item Many named curves; cubic bezier underneath.
  \item Spring physics for organic motion.
  \item Per-axis curves for 2-D.
  \item Cheap to evaluate.
  \item Respect reduced-motion preferences.
\end{itemize}
\end{recap}

\section*{Workshop}

\begin{enumerate}[leftmargin=*]
  \item Try linear, ease-out, spring on the same
    animation; pick the best.
  \item Animate a card with different curves per axis.
  \item Bypass curves when reduced motion is set.
\end{enumerate}
