\chapter{Progress, spinners, and indicators}
\label{ch:progress}

\begin{aside}
Long-running work needs to look like it's running. A bar that
fills, a spinner that turns, a percentage that climbs --- each
is a tiny widget with a big perceived impact.
\end{aside}

\section{Determinate progress}

A bar with known total:

\begin{lstlisting}
auto bar = progress_bar(percent)
         | width(30);
\end{lstlisting}

Renders as ``[\#\#\#\#\#.....]'' with hash for filled and
dot for unfilled. Subdivisions via half-block characters
for sub-cell precision.

\section{Unicode block fill}

For smooth bars, eight levels of fill per cell using
\code{U+2588} (full block) down to \code{U+258F} (left
eighth block):

\begin{lstlisting}
constexpr const char* fills[] = {
    " ", "\u258F", "\u258E", "\u258D",
    "\u258C", "\u258B", "\u258A", "\u2589",
    "\u2588"
};
\end{lstlisting}

A 30-cell bar at 33\% renders as 9 full blocks plus a
left-third block, then padding.

\section{Indeterminate spinner}

For unknown duration:

\begin{lstlisting}
constexpr const char* frames[] = {
    "|", "/", "-", "\\\\"
};

auto spinner = animated(100ms, [&] {
    return text(frames[ctx.frame_index() % 4]);
});
\end{lstlisting}

A 100 ms tick gives a 10 Hz spin --- snappy without being
seizure-inducing.

\section{Braille spinners}

A higher-resolution spinner uses braille dots
(\code{U+2800}--\code{U+28FF}). Each cell is an 8-dot grid;
animating which dots are lit gives 8-frame smoothness:

\begin{lstlisting}
constexpr const char* braille[] = {
    "\u2807", "\u280F", "\u281F", "\u287F",
    "\u287B", "\u2879", "\u2878", "\u2810"
};
\end{lstlisting}

Looks delightful.

\section{Multi-stage progress}

A job with phases (download, extract, install) shows the
current phase plus per-phase progress:

\begin{lstlisting}
auto multi = column(
    text("Downloading..."),
    progress_bar(download_pct),
    text("Extracting..."),
    progress_bar(extract_pct)
);
\end{lstlisting}

\section{Throughput display}

Beyond percent, show rate and ETA:

\begin{lstlisting}
auto status = text(
    format("{:.1f} MB/s, ETA {}",
        rate_mbps, eta)
);
\end{lstlisting}

\section{Silent failure}

A spinner that runs forever because the task crashed is the
worst UX. Wrap the task: on failure, switch to an error
state.

\section{Recap}

\begin{recap}
\begin{itemize}[leftmargin=*]
  \item Block characters for smooth bars.
  \item ASCII spinners for compatibility; braille for
    quality.
  \item Multi-stage for compound jobs.
  \item Rate and ETA add information.
  \item Handle failure visibly.
\end{itemize}
\end{recap}

\section*{Workshop}

\begin{enumerate}[leftmargin=*]
  \item Build a download widget with bar, rate, ETA.
  \item Create three spinner styles (ASCII, braille,
    custom). Pick one.
  \item Wire a failure path that shows the error and a
    retry button.
\end{enumerate}
