\chapter{Menus and command palettes}
\label{ch:menus}

\begin{aside}
A menu is a list with actions. A command palette is a menu
with search. Both are how power users navigate without
hunting for buttons.
\end{aside}

\section{Menu structure}

\begin{lstlisting}
struct MenuItem {
    std::string label;
    std::string shortcut;       // optional
    std::function<void()> action;
    bool separator;
    bool disabled;
};
\end{lstlisting}

A vertical list with the label on the left and the shortcut
on the right.

\section{Keyboard navigation}

Up/down moves the highlight; Enter activates. Letter
shortcuts (the first letter of each item) jump to the
matching item.

For nested menus, Right opens the submenu; Left closes.

\section{Command palette}

A modal overlay with a search box and a filtered list of
commands:

\begin{lstlisting}
auto palette = modal(column(
    text_input(query),
    virtual_list(
        filtered_commands(ctx.get(query)),
        [&](int i) { return command_row(...); }
    )
));
\end{lstlisting}

Trigger with Ctrl-P or Ctrl-Shift-P. Escape dismisses.

\section{Fuzzy matching}

The filter is fuzzy: ``op fl'' matches ``open file''. \maya{}
ships a fuzzy matcher (subset of the algorithm in
\code{fzf}).

\begin{lstlisting}
int fuzzy_score(std::string_view query, std::string_view target);
\end{lstlisting}

Higher score = better match. Sort by descending score.

\section{Recent commands}

A small history of recently invoked commands ranks them
higher in the palette. Persist across sessions.

\section{Argument prompts}

Some commands need arguments (open which file, rename to
what?). The palette can chain: pick command, then prompt
for argument.

\section{Context-aware commands}

The available commands depend on context: the current
view's verbs. Pass the context to the palette:

\begin{lstlisting}
auto commands = ctx.get(current_view).commands();
\end{lstlisting}

\section{Keyboard chords}

Multi-key bindings (\code{g d} for ``go to definition'') need
a chord buffer. After the first key, the runtime enters a
``waiting'' state; the next key resolves or times out.

\maya{}'s key binding system supports chords.

\section{Recap}

\begin{recap}
\begin{itemize}[leftmargin=*]
  \item Menu = list of items with actions.
  \item Palette = menu with fuzzy search.
  \item Recent commands ranked higher.
  \item Argument prompts via chaining.
  \item Chords for multi-key bindings.
\end{itemize}
\end{recap}

\section*{Workshop}

\begin{enumerate}[leftmargin=*]
  \item Build a command palette with 20 commands.
  \item Add fuzzy search and ranking by recency.
  \item Implement a chord binding for two-key sequences.
\end{enumerate}
