\chapter{Drag and drop}
\label{ch:dnd}

\begin{aside}
Drag-and-drop in a terminal feels modern. The implementation
is bookkeeping: pick-up on press, follow on move, drop on
release.
\end{aside}

\section{States}

\begin{itemize}[leftmargin=*]
  \item Idle.
  \item Dragging (carrying a payload).
  \item Hovering over a drop target.
  \item Drop in progress.
\end{itemize}

\section{Pick up}

On press over a draggable element, capture the payload:

\begin{lstlisting}
on_press([&] {
    ctx.set(drag_payload, item_id);
    ctx.set(drag_state, Drag::Active);
});
\end{lstlisting}

\section{Move}

On mouse motion, update the cursor position. Optionally
render a ghost element following the cursor.

\section{Drop targets}

Each potential target subscribes to drag events:

\begin{lstlisting}
on_drop([&](Payload p) {
    accept(p);
    ctx.set(drag_state, Drag::Idle);
});
\end{lstlisting}

\section{Drop preview}

While hovering a target, render a visual cue (highlight,
insertion indicator). Cancel on hover-out.

\section{Cancel}

Escape cancels the drag without dropping. Cursor returns,
state resets.

\section{Multi-select drag}

If multiple items are selected and one is dragged, the
whole selection drags. Payload is a list.

\section{Cross-window dnd}

Within a single \maya{} app, drag-and-drop is simple. Drag
to another terminal window is the OS's problem;
terminals don't propagate dnd events.

\section{Recap}

\begin{recap}
\begin{itemize}[leftmargin=*]
  \item Four states: idle, dragging, hovering, dropping.
  \item Ghost element follows cursor.
  \item Drop targets highlight on hover.
  \item Escape cancels.
  \item Multi-select drag carries a list.
\end{itemize}
\end{recap}

\section*{Workshop}

\begin{enumerate}[leftmargin=*]
  \item Build a drag-to-reorder list.
  \item Add a ghost element following the cursor.
  \item Implement escape-to-cancel.
\end{enumerate}
