\chapter{Dialogs and modals}
\label{ch:dialogs}

\begin{aside}
A dialog is a focused interaction that interrupts everything
else. Done well, it asks one question and gets out of the
way. Done badly, it stacks five deep and nobody knows where
they are.
\end{aside}

\section{The contract}

A dialog:

\begin{itemize}[leftmargin=*]
  \item Takes focus.
  \item Dims the background.
  \item Returns a result (yes/no/cancel, or a value).
  \item Closes on result or escape.
\end{itemize}

\section{The API}

\begin{lstlisting}
auto result = co_await ctx.dialog(
    confirm_dialog("Delete file?")
);
if (result == ConfirmResult::Yes) {
    delete_file();
}
\end{lstlisting}

The dialog is async --- the calling code awaits its result.

\section{Built-in dialogs}

\begin{itemize}[leftmargin=*]
  \item \code{confirm\_dialog(message)}: Yes/No.
  \item \code{alert\_dialog(message)}: OK.
  \item \code{prompt\_dialog(message, default)}: text input.
  \item \code{select\_dialog(options)}: pick one of many.
\end{itemize}

\section{Custom dialogs}

For complex flows, build your own:

\begin{lstlisting}
class MyDialog : public Dialog<MyResult> {
public:
    Element render() override { ... }
    void on_key(KeyEvent) override { ... }
};
\end{lstlisting}

The dialog framework handles focus and dismissal; your
class handles content and result.

\section{Stacking}

Dialogs can stack but it's a smell. \maya{} caps the stack
at 3 by default; further pushes warn.

\section{Dimming}

The background dims by halving the brightness of every
visible cell. \maya{} implements this in a post-paint pass
over the area outside the dialog.

\section{Escape key behaviour}

Escape on a confirm dialog returns ``cancel''. Escape on a
prompt returns nullopt. Configurable.

\section{Animations}

A dialog can animate in (fade, scale, slide). Subtle ones
work; flashy ones distract.

\section{Recap}

\begin{recap}
\begin{itemize}[leftmargin=*]
  \item Dialog = focused, async, returns a result.
  \item Built-in: confirm, alert, prompt, select.
  \item Custom via subclass.
  \item Dim the background.
  \item Cap stacking.
\end{itemize}
\end{recap}

\section*{Workshop}

\begin{enumerate}[leftmargin=*]
  \item Build a confirm-before-delete flow.
  \item Create a multi-step prompt (name, then email).
  \item Animate a dialog in and out.
\end{enumerate}
