\chapter{Search, find, and replace}
\label{ch:search}

\begin{aside}
A search box that filters in real time. Find next, find
previous, replace one, replace all. The pattern recurs in
every text-heavy TUI.
\end{aside}

\section{The search signal}

\begin{lstlisting}
auto query = signal<std::string>("");
auto matches = memo([&] {
    return find_all(content, ctx.get(query));
});
\end{lstlisting}

The memo re-runs only on query change.

\section{Case sensitivity}

Toggle case sensitivity via a checkbox or modifier:

\begin{lstlisting}
auto case_sensitive = signal<bool>(false);
\end{lstlisting}

Smart case (case-insensitive unless the query has
uppercase) is a common default.

\section{Regex}

For power users, allow regex queries. Toggle via flag.
\code{std::regex} works; \code{re2} is faster for large
corpora.

\section{Find next}

Walking through matches:

\begin{lstlisting}
auto current_match = signal(0);
on_key(KeyN, [&] {
    auto i = ctx.get(current_match);
    ctx.set(current_match, (i + 1) % matches.size());
});
\end{lstlisting}

\section{Visual highlighting}

Mark every match in the text with a highlighted
background. The current one gets a stronger highlight.

\section{Replace}

\begin{itemize}[leftmargin=*]
  \item Replace current.
  \item Replace all.
  \item Replace all in selection.
  \item Replace with regex capture groups.
\end{itemize}

\section{Incremental search}

Search updates on every keystroke. The first match becomes
the cursor position. Pressing Enter commits; Escape
cancels.

\section{Search history}

Up/down in the search box recalls previous queries. Persist
across sessions.

\section{Recap}

\begin{recap}
\begin{itemize}[leftmargin=*]
  \item Query signal + memoised matches.
  \item Smart case for friendliness.
  \item Regex for power.
  \item Highlight all + emphasis on current.
  \item Replace one/all with regex captures.
  \item Incremental, with history.
\end{itemize}
\end{recap}

\section*{Workshop}

\begin{enumerate}[leftmargin=*]
  \item Build an incremental search.
  \item Add regex support.
  \item Implement replace with capture groups.
\end{enumerate}
