\chapter{Notifications and toasts}
\label{ch:toasts}

\begin{aside}
A toast is a non-modal message that appears briefly and fades.
Notifications are the bigger sibling --- persistent, queued,
sometimes interactive.
\end{aside}

\section{Toast widget}

\begin{lstlisting}
ctx.toast("Saved!");
ctx.toast("Error", ToastLevel::Error);
\end{lstlisting}

The toast appears at a corner (top-right typically), holds
3 seconds, fades. Multiple toasts stack vertically.

\section{Levels}

\begin{itemize}[leftmargin=*]
  \item Info: blue or neutral.
  \item Success: green.
  \item Warning: yellow.
  \item Error: red.
\end{itemize}

\section{Persistence}

A toast vanishes on a timer. A notification stays until
dismissed. \maya{} treats them as separate types.

\section{Queue}

If toasts come faster than they expire, new ones queue
behind. \maya{}'s queue caps at 5 visible at a time;
overflow waits.

\section{Action toasts}

A toast can carry an action button:

\begin{lstlisting}
ctx.toast("File deleted", action("Undo", undo_fn));
\end{lstlisting}

The toast stays until clicked or timed out. Common for
``undo last action'' UX.

\section{Animation}

Slide in from the right, slide out to the right. 200 ms
each way. Respect reduced-motion.

\section{OS-level notifications}

For attention even when the terminal is unfocused, OSC 9
(see Chapter~\ref{ch:titles}) raises an OS notification.

\section{Recap}

\begin{recap}
\begin{itemize}[leftmargin=*]
  \item Toasts: transient, corner, fade.
  \item Levels colour-code severity.
  \item Action toasts include a button.
  \item OSC 9 for OS-level escalation.
\end{itemize}
\end{recap}

\section*{Workshop}

\begin{enumerate}[leftmargin=*]
  \item Wire toasts on save, delete, and error paths.
  \item Add an undo action toast after delete.
  \item Bridge to OS notifications for the error level.
\end{enumerate}
