\chapter{Auto-completion and suggestions}
\label{ch:autocomplete}

\begin{aside}
A text input that suggests as you type feels intelligent. The
implementation is filter plus dropdown plus keyboard
navigation.
\end{aside}

\section{The pattern}

\begin{lstlisting}
auto query = signal<std::string>("");
auto suggestions = memo([&] {
    return filter(corpus, ctx.get(query));
});
auto selected = signal(0);
\end{lstlisting}

The dropdown shows suggestions; up/down moves selected;
Tab or Enter commits.

\section{Filter strategies}

\begin{itemize}[leftmargin=*]
  \item Prefix match: starts-with.
  \item Substring: contains.
  \item Fuzzy: subset matching with scoring.
\end{itemize}

Fuzzy is the trend; prefix is the most predictable.

\section{Ranking}

Recency, frequency, exact-match-first. Personalise to the
user's history.

\section{Visual}

The dropdown appears below the input (or above if no room
below). Each row highlights the matching characters.

\section{Async sources}

Suggestions from a server arrive asynchronously. Debounce
the query so you don't fire on every keystroke; cancel
in-flight requests when the query changes.

\section{Commit and accept}

\begin{itemize}[leftmargin=*]
  \item Tab: accept current.
  \item Enter: accept and submit.
  \item Escape: dismiss.
  \item Up/Down: move selection.
\end{itemize}

\section{Multi-token}

For inputs with multiple tokens (file paths, email lists),
complete the current token only.

\section{Recap}

\begin{recap}
\begin{itemize}[leftmargin=*]
  \item Query + filtered suggestions + selected index.
  \item Fuzzy match with ranking.
  \item Async sources need debounce and cancel.
  \item Tab to accept, Escape to dismiss.
\end{itemize}
\end{recap}

\section*{Workshop}

\begin{enumerate}[leftmargin=*]
  \item Build an autocompleting search box.
  \item Add fuzzy matching.
  \item Wire to an async source with debounce.
\end{enumerate}
