\chapter{Layout splits and resizable panes}
\label{ch:splits}

\begin{aside}
A two-pane editor with a sidebar. A three-way diff with
gutters. Resizable splits give the user control over screen
real estate.
\end{aside}

\section{Split layouts}

A horizontal split divides cells left and right:

\begin{lstlisting}
auto split = hsplit(left_pane, right_pane, ratio);
\end{lstlisting}

Where \code{ratio} is the fraction allocated to the left
(0.0--1.0). 0.5 is even.

A vertical split is the same, top and bottom.

\section{Drag handles}

The boundary between panes is a draggable handle. \maya{}'s
split widget binds mouse events on the boundary cell:

\begin{lstlisting}
on_drag([&](int dx) {
    auto new_ratio = std::clamp(
        ratio + dx / float(total_width),
        0.1f, 0.9f
    );
    ctx.set(ratio_signal, new_ratio);
});
\end{lstlisting}

The clamp prevents collapsing one pane to nothing.

\section{Keyboard resize}

For accessibility, allow keyboard resize:

\begin{itemize}[leftmargin=*]
  \item Ctrl-Shift-Left/Right: shrink/grow the focused
    pane.
  \item Ctrl-Shift-=: reset to 50/50.
\end{itemize}

\section{Nested splits}

A horizontal split with a vertical split inside the right
pane is a three-pane layout:

\begin{lstlisting}
hsplit(
    left,
    vsplit(top_right, bottom_right, 0.5),
    0.3
);
\end{lstlisting}

\section{Snap}

Resize can snap to convenient ratios (1/3, 1/2, 2/3) when
near them. \maya{}'s split widget offers \code{snap\_to}
parameter.

\section{Collapsible panes}

A pane can collapse to zero width (hidden). Double-click
the handle to collapse; double-click again to restore.

The collapsed state remembers the previous ratio.

\section{Tabbed splits}

A split where one pane holds a tab strip is a
panes-with-tabs layout. Each tab has its own content;
switching tabs swaps the visible element.

\section{Persistence}

Save the ratio across sessions so the user's layout
persists. One signal per split, persisted to config.

\section{Recap}

\begin{recap}
\begin{itemize}[leftmargin=*]
  \item Splits divide layout by ratio.
  \item Drag handle for resize.
  \item Keyboard support for accessibility.
  \item Nesting builds N-pane layouts.
  \item Collapse, snap, persist for polish.
\end{itemize}
\end{recap}

\section*{Workshop}

\begin{enumerate}[leftmargin=*]
  \item Build a three-pane layout (sidebar, editor,
    preview).
  \item Add keyboard shortcuts for resize.
  \item Persist the ratios so the layout survives
    restart.
\end{enumerate}
