\chapter{Tabs, sections, and side rails}
\label{ch:tabs}

\begin{aside}
Tabs let one screen hold many views. The active tab fills the
content area; inactive tabs are labels at the top.
\end{aside}

\section{Tab strip}

\begin{lstlisting}
auto tabs = tab_strip({
    {"Home",     home_view()},
    {"Settings", settings_view()},
    {"About",    about_view()},
}, active);
\end{lstlisting}

The strip renders labels separated by some glyph; the
active one is highlighted.

\section{Keyboard navigation}

\begin{itemize}[leftmargin=*]
  \item Ctrl-Tab: next.
  \item Ctrl-Shift-Tab: previous.
  \item Ctrl-1..9: jump to tab N.
\end{itemize}

\section{Closable tabs}

A close glyph (x) per tab, click to close:

\begin{lstlisting}
auto label = row(text(name), close_button(on_close));
\end{lstlisting}

\section{Reorderable tabs}

Drag a tab left or right to reorder. The tab strip emits a
reorder event; the application updates the order signal.

\section{Vertical tabs}

A side rail is a vertical tab strip on the left edge. Icons
plus optional labels. Common for IDE-style tools.

\section{Tab groups}

Some apps group related tabs (file groups, workspaces).
Render group separators in the strip.

\section{Overflow}

Too many tabs? Two options:

\begin{itemize}[leftmargin=*]
  \item Scroll the strip horizontally.
  \item Collapse extras into a ``+5'' dropdown.
\end{itemize}

\maya{} defaults to scroll; dropdown is opt-in.

\section{Recap}

\begin{recap}
\begin{itemize}[leftmargin=*]
  \item Tab strip + content area.
  \item Keyboard nav with Ctrl-Tab.
  \item Closable, reorderable, vertical variants.
  \item Overflow via scroll or dropdown.
\end{itemize}
\end{recap}

\section*{Workshop}

\begin{enumerate}[leftmargin=*]
  \item Build a tabbed view with three tabs.
  \item Add Ctrl-W to close the current tab.
  \item Implement drag-to-reorder.
\end{enumerate}
