\chapter{Pickers: file, command, fuzzy}
\label{ch:pickers}

\begin{aside}
A picker is a modal list with a search box. File pickers,
command palettes, branch pickers, symbol pickers --- the
template is the same.
\end{aside}

\section{The components}

\begin{itemize}[leftmargin=*]
  \item Modal overlay.
  \item Search input at top.
  \item Scrollable list below.
  \item Highlighted current row.
  \item Optional preview pane on right.
\end{itemize}

\section{The data source}

\begin{lstlisting}
auto items = signal<std::vector<Item>>(...);
auto query = signal<std::string>("");
auto filtered = memo([&] {
    return fuzzy_filter(items, query);
});
\end{lstlisting}

\section{Async loading}

For large data (filesystem walk, git log), populate
asynchronously:

\begin{lstlisting}
spawn([&]{
    for (auto& path : walk_dir("/")) {
        ctx.post_to_main([path]{
            items.push_back(path);
        });
    }
});
\end{lstlisting}

The user sees results stream in.

\section{Preview pane}

The selected row drives a preview signal. The preview shows
file content, command help, diff, etc.

\section{Keybindings}

\begin{itemize}[leftmargin=*]
  \item Up/Down: move selection.
  \item Enter: open / execute.
  \item Tab: complete.
  \item Escape: dismiss.
  \item Ctrl-P / Ctrl-N: vim-like up/down.
\end{itemize}

\section{Multi-select}

For pickers that allow multiple selection (e.g., add files
to staging area), track a set:

\begin{lstlisting}
auto selected_set = signal<std::set<Id>>({});
\end{lstlisting}

Space toggles current; Enter commits the set.

\section{Recap}

\begin{recap}
\begin{itemize}[leftmargin=*]
  \item Modal + search + list + optional preview.
  \item Fuzzy filter via memo.
  \item Async stream for large sources.
  \item Multi-select with set + Space toggle.
\end{itemize}
\end{recap}

\section*{Workshop}

\begin{enumerate}[leftmargin=*]
  \item Build a file picker rooted at \code{\~/}.
  \item Add a preview pane.
  \item Implement multi-select with Tab.
\end{enumerate}
