\chapter{Diffs and merge views}
\label{ch:diffview}

\begin{aside}
Side-by-side diffs, unified diffs, three-way merges --- TUIs
shine here. Plain text + ANSI colour is exactly enough.
\end{aside}

\section{Diff representation}

\begin{lstlisting}
struct DiffLine {
    enum Kind { Same, Added, Removed, Modified };
    Kind kind;
    std::string text;
    int old_line, new_line;
};
\end{lstlisting}

Compute with Myers' algorithm or call out to \code{git
diff}.

\section{Unified diff view}

One column. Added lines green, removed red, context grey.
Line numbers in a gutter.

\section{Side-by-side}

Two columns. Old on left, new on right. Same lines align;
gaps fill with blanks where one side has no line.

\section{Inline diff}

Within a modified line, highlight added/removed words or
characters. Implement with a per-line word diff.

\section{Folding}

Long unchanged regions collapse to a placeholder:

\begin{lstlisting}
@@ collapsed 50 unchanged lines @@
\end{lstlisting}

Click or press a key to expand.

\section{Merge view}

Three columns for three-way merge: base, ours, theirs.
Conflicts show as overlapping changes.

\section{Navigation}

\begin{itemize}[leftmargin=*]
  \item Next/previous change.
  \item Jump to conflict.
  \item Stage chunk / unstage chunk.
\end{itemize}

\section{Recap}

\begin{recap}
\begin{itemize}[leftmargin=*]
  \item Diff lines with kind + text + line numbers.
  \item Unified vs side-by-side vs inline.
  \item Fold unchanged regions.
  \item Three-way for merges.
\end{itemize}
\end{recap}

\section*{Workshop}

\begin{enumerate}[leftmargin=*]
  \item Build a unified diff viewer for git output.
  \item Add side-by-side toggle.
  \item Implement next/prev change navigation.
\end{enumerate}
