\chapter{Notifications channel and pub/sub}
\label{ch:pubsub}

\begin{aside}
Some components need to talk to each other without sharing a
parent. A publish-subscribe channel lets one component emit
events the others observe, without coupling.
\end{aside}

\section{The channel}

\begin{lstlisting}
template<typename T>
class Channel {
    std::vector<std::function<void(const T&)>> subs;
public:
    void publish(const T& v) {
        for (auto& s : subs) s(v);
    }
    Subscription subscribe(auto fn) {
        subs.push_back(fn);
        return { ... };  // RAII unsubscribe
    }
};
\end{lstlisting}

\section{Typed channels}

A channel carries one payload type. Compile-time
type-checking prevents wrong-event delivery.

\begin{lstlisting}
Channel<FileChanged> file_channel;
Channel<UserLoggedIn> user_channel;
\end{lstlisting}

\section{Lifetime}

Subscriptions are RAII: the returned handle unsubscribes
on destruction. A component owns its subscription; when
the component unmounts, the subscription drops.

\section{Order of delivery}

Subscribers fire in subscription order. For ordering-
sensitive workflows, this matters.

\section{Async publish}

A publish that wants to defer to next frame uses a
microtask:

\begin{lstlisting}
ctx.microtask([&] { file_channel.publish(...); });
\end{lstlisting}

\section{Channels vs signals}

\begin{itemize}[leftmargin=*]
  \item \textbf{Signal:} stateful, latest value, eager
    notification.
  \item \textbf{Channel:} stateless, every event delivered,
    discrete.
\end{itemize}

Signals model state; channels model events.

\section{Global channels}

\maya{} provides built-in channels for common events:
window resize, focus gained/lost, theme changed. Subscribe
where useful.

\section{Recap}

\begin{recap}
\begin{itemize}[leftmargin=*]
  \item Typed pub/sub channels.
  \item RAII subscriptions.
  \item Async publish via microtask.
  \item Signals for state; channels for events.
\end{itemize}
\end{recap}

\section*{Workshop}

\begin{enumerate}[leftmargin=*]
  \item Build a file-changed channel; multiple components
    listen.
  \item Replace a global signal with a channel where
    appropriate.
  \item Subscribe to the built-in resize channel.
\end{enumerate}
