\chapter{File I/O and config files}
\label{ch:fileio}

\begin{aside}
A TUI typically reads a config file at startup and writes
state on exit. Both should be atomic, locale-stable, and
encoded in a forgiving format.
\end{aside}

\section{Where files live}

XDG Base Directory spec:

\begin{itemize}[leftmargin=*]
  \item Config: \code{\$XDG\_CONFIG\_HOME/myapp/config.toml}
    (default \code{\~/.config/myapp/}).
  \item Data: \code{\$XDG\_DATA\_HOME/myapp/} (default
    \code{\~/.local/share/myapp/}).
  \item Cache: \code{\$XDG\_CACHE\_HOME/myapp/} (default
    \code{\~/.cache/myapp/}).
\end{itemize}

\maya{} provides \code{xdg::config\_path("myapp")} that
expands correctly.

\section{Config format}

TOML, JSON, and YAML are the common choices. \maya{} ships
a TOML parser; users can write their own loaders for other
formats.

TOML is the recommended default: clean syntax, no
significant whitespace, comments, types.

\section{Loading on startup}

\begin{lstlisting}
auto cfg = load_config(xdg::config_path("myapp") + "/config.toml");
ctx.provide(cfg);
\end{lstlisting}

If the file doesn't exist, fall back to defaults. If it
exists but is malformed, log and use defaults.

\section{Atomic writes}

Writing to a file is not atomic if interrupted mid-write.
Use the rename trick:

\begin{lstlisting}
auto tmp = path + ".tmp";
std::ofstream(tmp) << content;
::rename(tmp.c_str(), path.c_str());
\end{lstlisting}

Rename is atomic on POSIX: the target either has the old
content or the new, never partial.

\section{Hot reload}

Watch the config file for changes:

\begin{lstlisting}
watch_file("config.toml", [&] {
    auto cfg = reload_config();
    ctx.set(config_signal, cfg);
});
\end{lstlisting}

The signal update propagates to all consumers.

\section{Locale and encoding}

Files are UTF-8 by default. Reading a non-UTF-8 file fails
with a clear error; don't silently re-interpret.

\section{File sizes}

A config file should be small (KB, not MB). For large data
(message history, cached search index), use a separate file
in the data directory.

\section{Recap}

\begin{recap}
\begin{itemize}[leftmargin=*]
  \item XDG paths for config, data, cache.
  \item TOML is the friendly format.
  \item Atomic writes via rename.
  \item Hot reload via file watcher.
  \item UTF-8 only; fail loud on others.
\end{itemize}
\end{recap}

\section*{Workshop}

\begin{enumerate}[leftmargin=*]
  \item Add a config file to your app. Load on startup.
  \item Wire atomic write on exit.
  \item Implement hot reload.
\end{enumerate}
