\chapter{Forms of state machines}
\label{ch:undostate}

\begin{aside}
Undo is a state machine: each action records a step; undo
rewinds. The data structure is a stack; the discipline is
that every mutation goes through the recorder.
\end{aside}

\section{The pattern}

\begin{lstlisting}
template<typename Action>
class History {
    std::vector<Action> undo_stack;
    std::vector<Action> redo_stack;
public:
    void apply(Action a) {
        a.do_();
        undo_stack.push_back(a);
        redo_stack.clear();  // new branch
    }
    void undo() { /* pop, undo_, push to redo */ }
    void redo() { /* pop, do_, push to undo */ }
};
\end{lstlisting}

\section{Action design}

Each action is reversible. Examples:

\begin{itemize}[leftmargin=*]
  \item Insert text at position N: undo by deleting N
    characters.
  \item Delete range: undo by re-inserting the saved text.
  \item Move cursor: undo by moving back.
\end{itemize}

\section{Grouping}

Multiple keystrokes into one undo unit. Group by:

\begin{itemize}[leftmargin=*]
  \item Time (within 500 ms of last).
  \item Boundary (space or punctuation starts new group).
\end{itemize}

\section{Branching}

When the user undoes then makes a new change, the redo stack
clears (the future is gone). Some editors implement a
branching history tree; \maya{} does not by default.

\section{Snapshot vs delta}

\begin{itemize}[leftmargin=*]
  \item Snapshot: store the full state. Memory-heavy.
  \item Delta: store the change. Smaller; needs reversal
    logic.
\end{itemize}

Delta is the standard for text editors. Snapshot is fine
for small data structures.

\section{Persistence}

Some apps persist undo history across sessions. Tricky:
the data must survive process restart and reattach to the
current state.

\section{Recap}

\begin{recap}
\begin{itemize}[leftmargin=*]
  \item Undo stack + redo stack.
  \item Each action reversible.
  \item Group by time or boundary.
  \item Delta is the standard.
\end{itemize}
\end{recap}

\section*{Workshop}

\begin{enumerate}[leftmargin=*]
  \item Add undo/redo to a text input.
  \item Implement grouping by 500 ms windows.
  \item Compare snapshot vs delta memory cost.
\end{enumerate}
