\chapter{Theming and skinning}
\label{ch:theming}

\begin{aside}
A theme is a colour palette plus a few style decisions. Two
apps with different themes look like cousins; same widgets,
different paint.
\end{aside}

\section{The theme struct}

\begin{lstlisting}
struct Theme {
    Color background, foreground;
    Color accent, muted;
    Color error, warning, success, info;
    Color border, focus;
    BorderChars borders;
    std::string name;
};
\end{lstlisting}

\section{Light and dark}

Two built-in themes:

\begin{itemize}[leftmargin=*]
  \item Light: bright background, dark text.
  \item Dark: dark background, bright text.
\end{itemize}

The user picks; the app honours.

\section{Detecting OS preference}

Some terminals report the OS theme via \code{COLORFGBG} or
DEC mode 2031. \maya{} reads on startup and picks a
matching theme.

\section{Custom themes}

Users define their own themes in config:

\begin{lstlisting}
[theme]
name = "solarized"
background = "#002b36"
accent = "#268bd2"
\end{lstlisting}

\section{Theme switching at runtime}

A theme signal:

\begin{lstlisting}
auto theme = signal<Theme>(load_default());
\end{lstlisting}

Effects on theme change re-render everything that consumes
it.

\section{Per-widget overrides}

A widget can ignore the theme for accent purposes:

\begin{lstlisting}
auto special = button("Critical")
             | fg<"#FF0000">;  // override
\end{lstlisting}

Use sparingly; consistency is a virtue.

\section{Accessible themes}

WCAG contrast ratios: 4.5:1 for body text, 3:1 for large
text. A theme that fails contrast is a UX bug.

\maya{}'s built-in themes meet AA contrast.

\section{Recap}

\begin{recap}
\begin{itemize}[leftmargin=*]
  \item Theme = palette + style decisions.
  \item Light and dark built-in; user-defined possible.
  \item OS theme detection on capable terminals.
  \item Theme signal for runtime switching.
  \item Mind contrast.
\end{itemize}
\end{recap}

\section*{Workshop}

\begin{enumerate}[leftmargin=*]
  \item Load a theme from config; switch at runtime.
  \item Detect dark/light mode from \code{COLORFGBG}.
  \item Measure contrast in your theme; fix anything
    below 4.5:1.
\end{enumerate}
