\chapter{Persistence and session restore}
\label{ch:persistence}

\begin{aside}
Closing your app and reopening should feel like you never
left. Open tabs, scroll positions, expanded folders, recent
files --- restore them all.
\end{aside}

\section{What to persist}

Two categories:

\begin{itemize}[leftmargin=*]
  \item \textbf{Preferences.} Theme, keymap, layout. User
    sets explicitly.
  \item \textbf{Session.} Open documents, cursor positions,
    scroll, undo history. Implicit.
\end{itemize}

Both go to disk; loads on startup.

\section{Format}

JSON or TOML for preferences (human-edited). Binary or
compressed JSON for session (machine-only).

\section{When to save}

\begin{itemize}[leftmargin=*]
  \item On exit (normal shutdown).
  \item On a timer (every 5 minutes).
  \item On crash (signal handler before re-raise).
  \item Never on every keystroke (too noisy).
\end{itemize}

\section{Versioning}

A schema version field lets the loader migrate older
formats:

\begin{lstlisting}
{ "version": 3, "tabs": [...], ... }
\end{lstlisting}

The loader reads the version, applies migrations, returns
the current shape.

\section{Atomic saves}

Write to a temp file, fsync, rename. The on-disk state is
always either old-complete or new-complete; never partial.

\section{Multiple windows}

If your app supports multiple instances, each needs its own
session file. Use the PID or a UUID.

\section{Encrypted persistence}

For sensitive state (passwords, tokens), encrypt with a
key from the OS keychain (libsecret on Linux, Keychain on
Mac).

\section{Recap}

\begin{recap}
\begin{itemize}[leftmargin=*]
  \item Preferences (human) vs session (machine).
  \item Save on exit, timer, and crash.
  \item Version the schema.
  \item Atomic rename for crash safety.
  \item OS keychain for secrets.
\end{itemize}
\end{recap}

\section*{Workshop}

\begin{enumerate}[leftmargin=*]
  \item Persist your app's open tabs.
  \item Add schema versioning and one migration.
  \item Implement timer-based autosave.
\end{enumerate}
