\chapter{Background sync and offline state}
\label{ch:offline}

\begin{aside}
A chat app should work on the train. A todo app should save
when the network drops. Offline-first design treats the
network as eventual, not required.
\end{aside}

\section{Local state first}

Every change writes to local storage immediately. The
UI reflects local state. Sync to the server is a
background concern.

\section{Sync queue}

Pending changes queue. When online, drain to the server in
order. When offline, the queue grows.

\section{Conflict resolution}

If two devices change the same record offline, sync brings
a conflict. Strategies:

\begin{itemize}[leftmargin=*]
  \item Last-write-wins (LWW): simple, lossy.
  \item Operational transform: complex, lossless.
  \item CRDT: complex, formally correct.
  \item Manual merge: user resolves.
\end{itemize}

\section{Visibility}

The user should see sync state:

\begin{itemize}[leftmargin=*]
  \item ``Synced'' or check mark when complete.
  \item ``Syncing...'' during a sync.
  \item ``Pending'' for queued items.
  \item ``Conflict'' for items needing resolution.
\end{itemize}

\section{Background fetch}

When online, periodically fetch updates from the server.
Merge with local state.

\section{Push notifications}

For instant updates, a WebSocket or server-sent events.
Falls back to polling when unavailable.

\section{Recap}

\begin{recap}
\begin{itemize}[leftmargin=*]
  \item Local first; sync second.
  \item Queue pending changes.
  \item Conflict resolution: LWW, OT, CRDT, manual.
  \item Show sync state in the UI.
\end{itemize}
\end{recap}

\section*{Workshop}

\begin{enumerate}[leftmargin=*]
  \item Implement local-first writes with a sync queue.
  \item Add sync state indicators.
  \item Try LWW conflict resolution; observe data loss
    scenarios.
\end{enumerate}
