\chapter{Selection and clipboard}
\label{ch:selection}

\begin{aside}
Selecting text and copying it should work in a TUI the way it
works everywhere else. Mouse drag, keyboard extend, copy on
release.
\end{aside}

\section{Selection types}

\begin{itemize}[leftmargin=*]
  \item Linear: start to end, one continuous span.
  \item Block: rectangular region, multi-column.
  \item Multi-cursor: many disjoint cursors.
\end{itemize}

\section{Mouse drag}

On mouse down, start a selection at the cell. On drag,
extend. On release, finalise. Visual feedback: invert the
selected cells.

\section{Keyboard extend}

Shift+arrow extends the selection by one cell. Shift+Home
extends to line start; Shift+End to line end.

\section{Word selection}

Double-click selects a word. Triple-click selects a line.

\section{Copy}

Selected text copies to the clipboard via OSC-52
(Chapter~\ref{ch:editors}). Some terminals require user
permission for OSC-52; allow but expect failure.

\section{Cut}

Cut is copy plus delete. Same as copy if the source is
read-only.

\section{Paste}

Bracketed paste delivers the pasted text as one event. The
focused element inserts at the cursor.

\section{Primary selection (X11)}

Primary selection on Linux: middle-click pastes the last
selection. OSC-52 \code{p} target supports this.

\section{Recap}

\begin{recap}
\begin{itemize}[leftmargin=*]
  \item Linear, block, multi-cursor selections.
  \item Mouse drag + keyboard extend.
  \item Double/triple click for word/line.
  \item OSC-52 for copy; bracketed paste for paste.
  \item Primary selection on X11.
\end{itemize}
\end{recap}

\section*{Workshop}

\begin{enumerate}[leftmargin=*]
  \item Implement linear selection with mouse drag.
  \item Add word and line selection.
  \item Wire OSC-52 copy and verify on your terminal.
\end{enumerate}
