\chapter{The signal-graph debugger}
\label{ch:signaldbg}

\begin{aside}
A regression like ``why does this widget re-render every
frame'' is solved by examining the dependency graph. The
signal debugger is the tool.
\end{aside}

\section{Inspecting signals}

\begin{lstlisting}
ctx.dump_signals();
// signal:user_id (ver 5) subs: 3
// signal:theme (ver 1) subs: 12
// ...
\end{lstlisting}

The dump lists every signal with version and subscriber
count.

\section{Tracing updates}

A trace flag records every set:

\begin{lstlisting}
ctx.set_trace_signals(true);
\end{lstlisting}

Each \code{set} logs to the trace file with stack trace.
The log shows ``who set what when''.

\section{Visualising the graph}

\code{ctx.dump\_graph(path)} writes graphviz dot format.
Render to PNG with \code{dot -Tpng}; eyeball the graph.

\section{Find re-render culprits}

When something re-renders too often, find the signal that
ticks too often:

\begin{itemize}[leftmargin=*]
  \item Filter signals by update rate.
  \item Inspect their subscribers.
  \item Identify which effect is wasteful.
\end{itemize}

\section{Replay}

Record an input sequence; replay with the debugger
attached. Step through the signal updates frame by frame.

\section{Breakpoints}

Set a breakpoint on a signal update:

\begin{lstlisting}
ctx.break_on_set(user_id);
\end{lstlisting}

The next set fires the debugger.

\section{Recap}

\begin{recap}
\begin{itemize}[leftmargin=*]
  \item Dump signals to see versions and sub counts.
  \item Trace updates to find culprits.
  \item Visualise the graph with graphviz.
  \item Breakpoints on specific signals.
\end{itemize}
\end{recap}

\section*{Workshop}

\begin{enumerate}[leftmargin=*]
  \item Dump your app's signal graph.
  \item Find a wasteful re-render path.
  \item Set a breakpoint on a critical signal.
\end{enumerate}
