\chapter{Patterns: model, view, presenter}
\label{ch:mvp}

\begin{aside}
Separation of concerns: model holds data, view renders, the
glue between them is the presenter (or controller, or view
model). The names vary; the idea persists.
\end{aside}

\section{The three layers}

\begin{itemize}[leftmargin=*]
  \item \textbf{Model.} Pure data and business logic. No
    \maya{} types.
  \item \textbf{View.} Pure rendering. Takes data, returns
    elements.
  \item \textbf{Presenter.} The glue: signal bindings,
    event handlers.
\end{itemize}

\section{The view}

\begin{lstlisting}
Element user_view(const User& u) {
    return column(
        text(u.name),
        text(u.email)
    );
}
\end{lstlisting}

No signals, no callbacks. Just data in, element out. Easy
to test.

\section{The presenter}

\begin{lstlisting}
Element user_presenter() {
    auto u = ctx.consume<UserStore>();
    return user_view(ctx.get(u->user));
}
\end{lstlisting}

The presenter wires the view to the store.

\section{Why this matters}

\begin{itemize}[leftmargin=*]
  \item Views are testable without a context.
  \item Models are testable without UI.
  \item Presenters are thin and obvious.
\end{itemize}

\section{Other patterns}

MVVM (Model-View-ViewModel) is similar; the ViewModel
holds the bindings. MVC (Controller) is older and similar
shape. Pick whichever vocabulary your team prefers.

\section{Anti-patterns}

\begin{itemize}[leftmargin=*]
  \item God presenter: knows everything, fragile.
  \item View with signals: tightly coupled, hard to test.
  \item Model with UI: pollutes the data layer.
\end{itemize}

\section{Recap}

\begin{recap}
\begin{itemize}[leftmargin=*]
  \item Model: data. View: render. Presenter: glue.
  \item Views are pure functions.
  \item Models have no UI types.
  \item Presenters are thin.
\end{itemize}
\end{recap}

\section*{Workshop}

\begin{enumerate}[leftmargin=*]
  \item Refactor one screen into model/view/presenter.
  \item Write a unit test for the view (no context).
  \item Identify any god presenter in your code; split
    it.
\end{enumerate}
