\chapter{Concurrency patterns}
\label{ch:concurrency}

\begin{aside}
A TUI is mostly single-threaded, but the I/O around it is
not. Talking to disks, networks, and computations off the
main thread needs patterns.
\end{aside}

\section{Single-thread UI}

\maya{}'s render loop runs on one thread. All signal
access is from this thread.

\section{Worker pool for CPU}

CPU-bound work goes to a pool. Results come back via
\code{post\_to\_main}.

\section{Reactor for I/O}

I/O-bound work goes to a reactor (\code{poll},
\code{epoll}, \code{kqueue}). Each fd has a callback that
runs on the main thread when readable.

\section{Channels for communication}

A multi-producer single-consumer queue is the bridge.
Workers push results; main pops in batches.

\section{Locking}

For shared mutable state across threads, use a mutex. Hold
it for short durations. Never hold a mutex across an
\code{await}.

\section{Async coroutines}

Coroutines structure async work as if it were sync. The
suspend points are explicit (\code{co\_await}).

\section{Anti-patterns}

\begin{itemize}[leftmargin=*]
  \item Touching signals from a worker thread.
  \item Holding a lock during render.
  \item Blocking the main thread for I/O.
  \item Spawning a thread per task (use the pool).
\end{itemize}

\section{Recap}

\begin{recap}
\begin{itemize}[leftmargin=*]
  \item Main thread for UI.
  \item Pool for CPU work.
  \item Reactor for I/O.
  \item Channels and post\_to\_main as bridges.
  \item Mutex sparingly; never across await.
\end{itemize}
\end{recap}

\section*{Workshop}

\begin{enumerate}[leftmargin=*]
  \item Move a CPU task to the pool; post result back.
  \item Add a reactor watcher for a socket.
  \item Find and fix any worker-thread signal access.
\end{enumerate}
