\chapter{Building a file manager}
\label{ch:filemgr}

\begin{aside}
Two-pane file managers (mc, ranger). Navigate with arrows,
preview on the right. Familiar workflow.
\end{aside}

\section{Layout}

Two columns. Left is the current directory list. Right is
a preview of the selected file.

\section{Directory listing}

Read the directory; for each entry produce a row with
name, size, type, modified time.

\begin{lstlisting}
std::vector<DirEntry> list_dir(const std::filesystem::path& p);
\end{lstlisting}

Sort by name (or by other column on user request).

\section{Selection}

A signal holds the selected index. Arrows move; Enter
opens (cd into directory or open file).

\section{Preview}

The preview shows:

\begin{itemize}[leftmargin=*]
  \item Text files: first 1000 lines, syntax-highlighted.
  \item Images: rendered via image protocol.
  \item Binaries: file type and size.
  \item Directories: child entry count.
\end{itemize}

\section{Navigation}

\begin{itemize}[leftmargin=*]
  \item Up/down: select.
  \item Enter: open / cd.
  \item Backspace: cd ..
  \item / : focus search box.
  \item D : delete (with confirm).
  \item R : rename.
  \item Y : copy.
  \item X : cut.
  \item P : paste.
\end{itemize}

\section{Bookmarks}

User-defined paths quick-jump to: \code{Ctrl-1} through
\code{Ctrl-9}. Persisted in config.

\section{Hidden files}

Toggle with \code{.}.

\section{Recap}

\begin{recap}
\begin{itemize}[leftmargin=*]
  \item Two panes: list + preview.
  \item Arrows + Enter for navigation.
  \item Search, delete, rename, copy/cut/paste.
  \item Bookmarks for jumps.
\end{itemize}
\end{recap}

\section*{Workshop}

\begin{enumerate}[leftmargin=*]
  \item Build the two-pane shell.
  \item Implement preview for text and images.
  \item Add bookmarks and the hidden-files toggle.
\end{enumerate}
