\chapter{Anti-patterns to avoid}
\label{ch:antipatterns}

\begin{aside}
Things that look reasonable but are wrong. Walk the path
once; learn to avoid the next time.
\end{aside}

\section{Reading signals in render order}

If component A's render reads a signal that B's render
sets, you have a race. Render the producer first.

\section{Mutating during render}

A render function should be pure. Setting a signal during
render schedules a re-render --- infinite loop.

\section{Background thread touching signals}

Signals are not thread-safe. Always post results back to
the main thread.

\section{Synchronous I/O on main thread}

A blocking read freezes the UI. Always async.

\section{Long-running effects}

An effect should compute quickly. Long work goes to a
worker thread.

\section{Too many effects}

An effect per item in a 10,000-item list is wasteful. Use
a single effect that handles all items.

\section{Reaching into widget internals}

Widgets are opaque. If you find yourself reaching in,
either compose differently or extend the widget's API.

\section{Hardcoded colours}

Theme-aware code uses theme tokens, not literal hex. Hex
literals are for the theme definition itself.

\section{Hardcoded sizes}

Layout systems exist; let them work. Hardcoded pixel
sizes (cell sizes) are fragile across terminal sizes.

\section{No keyboard path}

Mouse-only widgets exclude users. Always provide a
keyboard alternative.

\section{Ignoring NO\_COLOR}

Some users need colour off. Respect.

\section{Crashing on resize}

Resize is common. Test by dragging the window.

\section{Allocating per frame}

Allocations add up. Use the arena; reuse buffers.

\section{Recap}

\begin{recap}
\begin{itemize}[leftmargin=*]
  \item Pure render; no side effects.
  \item Signals are single-threaded.
  \item Async I/O always.
  \item Theme tokens, not hex.
  \item Keyboard parity with mouse.
  \item Respect NO\_COLOR.
\end{itemize}
\end{recap}

\section*{Workshop}

\begin{enumerate}[leftmargin=*]
  \item Audit your code for any anti-pattern above.
  \item Fix the worst offender.
  \item Add a lint or grep to catch future occurrences.
\end{enumerate}
