\chapter{Future of TUI: graphics protocols}
\label{ch:future_graphics}

\begin{aside}
The terminal has been a character grid for fifty years. The
last five have seen it learn to render images, true colour,
hyperlinks, and synchronised frames. The next five will see
even more.
\end{aside}

\section{Where we are}

\begin{itemize}[leftmargin=*]
  \item Truecolour: now universal.
  \item OSC-8 links: most modern terminals.
  \item Image protocols: three competing, narrowing to two.
  \item DEC 2026: arrived.
  \item CSI-u keyboard: arrived.
\end{itemize}

\section{Where it's going}

\begin{itemize}[leftmargin=*]
  \item Animation primitives (terminal-side tweens).
  \item Vector graphics protocol.
  \item Standardised image protocol (sixel and Kitty
    converge or one wins).
  \item Native multi-window in terminals.
  \item GPU shaders for terminal post-processing.
\end{itemize}

\section{What's stuck}

\begin{itemize}[leftmargin=*]
  \item Apple Terminal: no new features in years.
  \item tmux passthrough: better than nothing but lossy.
  \item BiDi: unsolved at the protocol level.
\end{itemize}

\section{What \maya{} will do}

Track the new features as they ship. The internal model
already separates capability detection from emission;
adding new outputs is a matter of feature flags.

\section{The big speculation}

Will the terminal become a GUI? Probably not in the
near term. The character grid is too useful, too fast,
too SSH-friendly. Innovations layer on top, not replace.

\section{Recap}

\begin{recap}
\begin{itemize}[leftmargin=*]
  \item Truecolour, links, images, synced output now
    common.
  \item More animation and graphics likely soon.
  \item Apple Terminal and tmux are the stuck links.
  \item The grid is here to stay.
\end{itemize}
\end{recap}

\section*{Workshop}

\begin{enumerate}[leftmargin=*]
  \item Check which features your daily terminal
    supports.
  \item Try Kitty or Wezterm; observe the difference.
  \item Speculate: what feature do you most want next?
\end{enumerate}
