\chapter{Future of TUI: input}
\label{ch:future_input}

\begin{aside}
Input has lagged output. The keyboard sends bytes
ambiguously; the mouse lost half its bits to compatibility.
CSI-u and the Kitty protocol promise better.
\end{aside}

\section{What CSI-u solves}

\begin{itemize}[leftmargin=*]
  \item Unique codes per key.
  \item Modifier-only events (shift held alone).
  \item Distinct press vs release.
  \item Repeat events.
\end{itemize}

\section{What's still missing}

\begin{itemize}[leftmargin=*]
  \item Touch gestures (pinch, rotate).
  \item Tablet/stylus input.
  \item Voice or transcription input.
  \item Eye tracking.
\end{itemize}

These are unlikely to land in terminal protocols soon ---
they belong to the OS, not the terminal.

\section{Multi-modal future}

A terminal that supports clicks plus keyboard plus
graphics could host much richer UIs than today's. The
limit is what protocols the community agrees on.

\section{Compose and IME}

Current IME flows are good. CJK input works. Improvements
here would be incremental.

\section{Accessibility}

Screen reader integration in terminals is poor. A
standardised ``announce this'' protocol would help.

\section{What \maya{} can do}

When new input protocols land, expose them. The internal
event types are open for extension.

\section{Recap}

\begin{recap}
\begin{itemize}[leftmargin=*]
  \item CSI-u solves keyboard ambiguity.
  \item Touch and stylus unlikely to come to terminals
    soon.
  \item A11y in terminals is the next frontier.
\end{itemize}
\end{recap}

\section*{Workshop}

\begin{enumerate}[leftmargin=*]
  \item Test CSI-u in your terminal; identify
    untrackable keys.
  \item Try a terminal screen reader; observe gaps.
\end{enumerate}
