\chapter{Future of TUI: the library}
\label{ch:future_library}

\begin{aside}
What's next for \maya{}? A list of things in progress,
planned, or wished for.
\end{aside}

\section{In progress}

\begin{itemize}[leftmargin=*]
  \item BiDi support via libfribidi.
  \item Plug-in ABI v2.
  \item Native widget themes (matching popular editors).
  \item Web export (canvas-based).
\end{itemize}

\section{Planned}

\begin{itemize}[leftmargin=*]
  \item Hot reload built into the runtime.
  \item Snapshot test framework.
  \item Benchmark suite expansion.
  \item More widget catalogue (date pickers, charts).
\end{itemize}

\section{Wished for}

\begin{itemize}[leftmargin=*]
  \item Mobile terminal support (iSH, Termux native).
  \item GUI backend (render to a window without
    terminal).
  \item AOT compilation of element trees.
\end{itemize}

\section{What's stable}

The core: signal graph, element tree, diff loop. Changes
here are rare and additive.

\section{What's evolving}

The widget catalogue: more widgets, refined APIs.

\section{Versioning}

Semver. Major bumps for breaking changes. The book tracks
the latest.

\section{Community}

Issues, PRs, discussions. See Chapter~\ref{ch:contributing}.

\section{Recap}

\begin{recap}
\begin{itemize}[leftmargin=*]
  \item Core stable, widgets evolving.
  \item Hot reload, snapshot framework planned.
  \item Mobile and GUI backends wished.
\end{itemize}
\end{recap}

\section*{Workshop}

\begin{enumerate}[leftmargin=*]
  \item File an issue suggesting a feature.
  \item Look at the planned list; pick one to implement.
\end{enumerate}
