\chapter{Common bugs and their fixes}
\label{ch:commonbugs}

\begin{aside}
The bugs you'll hit at least once. Listed by symptom, with
the fix.
\end{aside}

\section{Symptom: screen freezes after some input}

\textbf{Cause:} The event loop is blocked.

\textbf{Fix:} Find the synchronous I/O on main thread.
Move to async.

\section{Symptom: widget loses focus mysteriously}

\textbf{Cause:} Tree rebuild without keys for list items.

\textbf{Fix:} Add stable keys to list rows.

\section{Symptom: signal updates don't trigger re-render}

\textbf{Cause:} The signal was read outside an effect, so
no subscription.

\textbf{Fix:} Move the read into an effect or memo.

\section{Symptom: 100\% CPU on idle}

\textbf{Cause:} A clock signal subscribed to by an effect
that writes a signal read by the same effect.

\textbf{Fix:} Break the cycle. Don't depend on the clock
from a non-animation effect.

\section{Symptom: terminal stuck after crash}

\textbf{Cause:} Atexit hook didn't run.

\textbf{Fix:} Run \code{reset} in the shell. Investigate
why atexit was skipped (signal-killed?).

\section{Symptom: garbled text in mixed CJK content}

\textbf{Cause:} Width table mismatch with the terminal.

\textbf{Fix:} Set ambiguous-width to match the terminal
(usually 1; CJK locale users want 2).

\section{Symptom: missed keystrokes during paste}

\textbf{Cause:} Bracketed paste not enabled.

\textbf{Fix:} \maya{} enables it by default; verify
nothing disabled it.

\section{Symptom: animations jerky}

\textbf{Cause:} Long frames; or render-on-tick fallacy.

\textbf{Fix:} Profile. If renders take >16 ms, optimise.
If renders fire too often, debounce.

\section{Symptom: ANSI bytes leaking}

\textbf{Cause:} Untrusted string with control bytes
emitted directly.

\textbf{Fix:} Sanitise before display
(Chapter~\ref{ch:security2}).

\section{Recap}

\begin{recap}
\begin{itemize}[leftmargin=*]
  \item Blocked loop = sync I/O on main.
  \item Lost focus = missing keys.
  \item No re-render = read outside effect.
  \item Idle CPU = signal cycle.
\end{itemize}
\end{recap}

\section*{Workshop}

\begin{enumerate}[leftmargin=*]
  \item Pick a symptom you've hit; verify the cause.
  \item Add a regression test.
\end{enumerate}
