\chapter{Performance budgets per widget}
\label{ch:perfbudget}

\begin{aside}
A reference table: how much each widget costs at typical
sizes. Use to estimate your app's frame budget.
\end{aside}

\section{Measurement method}

Each widget measured in isolation: render to a string
backend, time 1000 iterations, divide. Numbers from a
mid-range laptop (M1 / Ryzen 7).

\section{Layout cost}

\begin{itemize}[leftmargin=*]
  \item Empty column / row: <1 microsecond.
  \item 100-element list: ~50 microseconds.
  \item 1000-element list (no virtualisation): ~500 us.
  \item Nested grid with 50 cells: ~30 us.
  \item Virtualised list (50 visible of 10000): ~30 us.
\end{itemize}

\section{Paint cost}

\begin{itemize}[leftmargin=*]
  \item Solid background fill: cells / 100 us.
  \item Text run: chars / 200 us.
  \item Border: corners + sides; <5 us.
  \item Image (cached): <10 us.
  \item Gradient: cells / 50 us.
\end{itemize}

\section{Diff cost}

\begin{itemize}[leftmargin=*]
  \item Unchanged 80x24 grid (SIMD): <50 us.
  \item Fully changed 80x24 grid: ~200 us emit + 50 us diff.
  \item Typical small update: <10 us.
\end{itemize}

\section{Signal cost}

\begin{itemize}[leftmargin=*]
  \item Read: ~10 ns.
  \item Write with 0 subscribers: ~30 ns.
  \item Write with 10 subscribers: ~200 ns.
  \item Memo recomputation: depends on function.
\end{itemize}

\section{Frame budget}

For 60 Hz: 16 ms. For 30 Hz: 33 ms. Most apps comfortably
fit in 5 ms of layout + paint + diff + emit.

\section{Big numbers}

To exceed budget on a typical laptop you need either:

\begin{itemize}[leftmargin=*]
  \item A 500x150 fully-changing grid.
  \item Tens of thousands of unbatched signal updates.
  \item Deep nesting (>1000 layers).
\end{itemize}

\section{Recap}

\begin{recap}
\begin{itemize}[leftmargin=*]
  \item Most operations: microseconds.
  \item Budget: 5 ms of the 16 ms frame.
  \item Pathological cases exist but are rare.
\end{itemize}
\end{recap}

\section*{Workshop}

\begin{enumerate}[leftmargin=*]
  \item Measure each widget in your app.
  \item Identify the most expensive.
  \item Optimise or virtualise.
\end{enumerate}
