\chapter{Patterns: portal rendering}
\label{ch:portals}

\begin{aside}
Sometimes a component's lexical position differs from its
visual position. A tooltip lives next to its anchor in the
code but renders above everything in the visual layer.
\end{aside}

\section{Portal mechanics}

A portal element registers its subtree with the renderer
to paint at a different time:

\begin{lstlisting}
auto popup = portal(
    panel("Help") | absolute(10, 5)
);
\end{lstlisting}

The popup lives in the tree where written but paints in
the overlay layer.

\section{Multiple portal targets}

\maya{} has named layers:

\begin{itemize}[leftmargin=*]
  \item \code{main}: the default.
  \item \code{overlay}: above main.
  \item \code{modal}: above overlay.
  \item \code{tooltip}: above modal.
\end{itemize}

\code{portal\_to("modal", element)} targets a specific
layer.

\section{Z-ordering within layers}

Within a layer, paint order is registration order. Last
registered paints on top.

\section{Use cases}

\begin{itemize}[leftmargin=*]
  \item Tooltips near anchors.
  \item Dropdowns from inputs.
  \item Floating action buttons.
  \item Status overlays.
\end{itemize}

\section{Recap}

\begin{recap}
\begin{itemize}[leftmargin=*]
  \item Portals decouple lexical and visual position.
  \item Named layers stack.
  \item Last registered paints on top within a layer.
\end{itemize}
\end{recap}

\section*{Workshop}

\begin{enumerate}[leftmargin=*]
  \item Build a tooltip via portal.
  \item Compare with inline rendering; observe Z issues.
  \item Use multiple layers in a complex screen.
\end{enumerate}
