\chapter{Patterns: graceful degradation}
\label{ch:degradation}

\begin{aside}
The user's terminal may not support what you assume. Detect,
degrade, continue.
\end{aside}

\section{What to detect}

\begin{itemize}[leftmargin=*]
  \item Colour mode (16, 256, truecolour).
  \item Image protocol (Sixel, Kitty, iTerm2, none).
  \item OSC-8 hyperlinks.
  \item CSI-u keyboard.
  \item DEC 2026 sync.
  \item Mouse modes.
  \item Bracketed paste.
\end{itemize}

\section{Fallbacks}

For each capability, a fallback:

\begin{itemize}[leftmargin=*]
  \item Truecolour \textrightarrow{} 256 \textrightarrow{}
    16.
  \item Image \textrightarrow{} half-block image
    \textrightarrow{} ASCII art \textrightarrow{} text
    label.
  \item Hyperlink \textrightarrow{} plain text (URL shown).
  \item CSI-u \textrightarrow{} legacy keys (some
    distinctions lost).
\end{itemize}

\section{Detection at startup}

Query the terminal; parse the reply. Cache results for the
session.

\section{Runtime checks}

\begin{lstlisting}
if (ctx.caps().has_image) {
    return image(path);
} else {
    return block_image(path);
}
\end{lstlisting}

\section{Manual override}

A config flag lets users force a specific level (testing,
or terminals that lie about capabilities).

\section{Recap}

\begin{recap}
\begin{itemize}[leftmargin=*]
  \item Detect capabilities at startup.
  \item Provide fallbacks at every level.
  \item Allow manual override.
\end{itemize}
\end{recap}

\section*{Workshop}

\begin{enumerate}[leftmargin=*]
  \item Test your app in xterm-16, xterm-256, kitty.
  \item Add fallbacks where missing.
  \item Add a config override flag.
\end{enumerate}
