\chapter{Patterns: graceful exit}
\label{ch:exit}

\begin{aside}
The user wants to quit. Save unsaved work. Confirm
destructive intent. Restore the terminal. Then exit.
\end{aside}

\section{Exit triggers}

\begin{itemize}[leftmargin=*]
  \item Quit command (Ctrl-Q, :q, menu File>Quit).
  \item Window close.
  \item SIGTERM, SIGINT.
  \item Crash.
\end{itemize}

\section{Unsaved work}

If state is dirty, prompt: ``Save changes? [yes/no/cancel]''.
Cancel aborts the exit.

\section{Shutdown sequence}

\begin{enumerate}[leftmargin=*]
  \item Stop accepting input.
  \item Save persistent state (atomic).
  \item Disconnect from external services.
  \item Cancel background tasks.
  \item Restore terminal (alt-screen exit, modes off).
\end{enumerate}

\section{Crash handler}

A SIGSEGV handler that restores the terminal before
re-raising:

\begin{lstlisting}
std::signal(SIGSEGV, [](int sig) {
    backend->shutdown();
    std::signal(sig, SIG_DFL);
    std::raise(sig);
});
\end{lstlisting}

\section{Atexit}

Belt and suspenders:

\begin{lstlisting}
std::atexit([] { backend.shutdown(); });
\end{lstlisting}

Idempotent so multiple invocations are safe.

\section{Recap}

\begin{recap}
\begin{itemize}[leftmargin=*]
  \item Confirm unsaved work.
  \item Atomic save before exit.
  \item Restore terminal in all paths.
  \item Crash handler + atexit.
\end{itemize}
\end{recap}

\section*{Workshop}

\begin{enumerate}[leftmargin=*]
  \item Add an unsaved-work prompt to your quit command.
  \item Install crash + atexit handlers.
  \item Test with kill -9 (won't help) and kill -TERM
    (should clean up).
\end{enumerate}
