\chapter{Appendix: troubleshooting}
\label{ch:troubleshoot}

\begin{aside}
Common things that go wrong, in checklist form.
\end{aside}

\section{App won't build}

\begin{itemize}[leftmargin=*]
  \item Compiler too old. Need GCC 14+ or Clang 18+.
  \item \code{-std=c++26} or \code{-std=c++2c} missing.
  \item \code{add\_subdirectory(maya)} forgotten.
  \item Linker can't find \code{maya} target ---
    check \code{target\_link\_libraries}.
\end{itemize}

\section{Screen looks wrong}

\begin{itemize}[leftmargin=*]
  \item Wrong ambiguous-width setting.
  \item Terminal font missing glyphs.
  \item Truecolour requested on a 16-colour terminal.
  \item Wide cells not handled (font shows half).
\end{itemize}

\section{App freezes}

\begin{itemize}[leftmargin=*]
  \item Synchronous I/O on main thread.
  \item Infinite loop in an effect.
  \item Deadlock with worker thread.
\end{itemize}

\section{Memory grows}

\begin{itemize}[leftmargin=*]
  \item Effect leak: not unmounted.
  \item Detached effect retained.
  \item Cache without eviction.
  \item Long-lived strings in arena (must be trivially
    destructible).
\end{itemize}

\section{Crashes on resize}

\begin{itemize}[leftmargin=*]
  \item Old size cached and stale.
  \item Element math overflows on weird sizes (0 or 1
    cell).
  \item Layout assumes minimum size.
\end{itemize}

\section{Keys not received}

\begin{itemize}[leftmargin=*]
  \item Raw mode not enabled.
  \item Wrong terminal --- some emit different escapes.
  \item CSI-u disabled by tmux.
  \item Focus on wrong element.
\end{itemize}

\section{Mouse not working}

\begin{itemize}[leftmargin=*]
  \item Mouse mode disabled.
  \item Terminal doesn't support SGR mode.
  \item tmux without mouse passthrough.
\end{itemize}

\section{Colours wrong}

\begin{itemize}[leftmargin=*]
  \item \code{NO\_COLOR} set.
  \item Wrong COLORTERM.
  \item Theme not loaded.
\end{itemize}

\section{Recap}

\begin{recap}
\begin{itemize}[leftmargin=*]
  \item Build: compiler, flags, link.
  \item Display: width, font, terminal caps.
  \item Behaviour: threads, effects, lifecycles.
\end{itemize}
\end{recap}
