\chapter{Appendix: parting checklist}
\label{ch:partingchecklist}

\begin{aside}
Before you ship, before you announce, before you call it
done. One last pass over the obvious.
\end{aside}

\section{Functionality}

\begin{itemize}[leftmargin=*]
  \item Every advertised feature works.
  \item Defaults make sense for first run.
  \item Empty states are designed.
  \item Errors are visible and actionable.
\end{itemize}

\section{Performance}

\begin{itemize}[leftmargin=*]
  \item Idle CPU is near zero.
  \item Startup completes under one second.
  \item Typing latency under twenty milliseconds.
  \item No memory leaks under one-hour soak.
\end{itemize}

\section{Polish}

\begin{itemize}[leftmargin=*]
  \item Animations respect reduced-motion.
  \item Colours respect \code{NO\_COLOR}.
  \item Theme contrast meets AA.
  \item Cursor restored on exit.
\end{itemize}

\section{Robustness}

\begin{itemize}[leftmargin=*]
  \item Survives resize during typing.
  \item Survives Ctrl-Z and resume.
  \item Survives SIGTERM with cleanup.
  \item Crash leaves terminal usable.
\end{itemize}

\section{Distribution}

\begin{itemize}[leftmargin=*]
  \item Static or properly-linked binary.
  \item Signed and checksummed.
  \item README explains install in 5 lines.
  \item Changelog covers since last release.
\end{itemize}

\section{Documentation}

\begin{itemize}[leftmargin=*]
  \item Help screen lists every key.
  \item Man page or web docs for reference.
  \item Asciinema cast for quick visual.
  \item Issues link in --help output.
\end{itemize}

\section{Recap}

\begin{recap}
\begin{itemize}[leftmargin=*]
  \item Functionality, performance, polish, robustness,
    distribution, documentation.
  \item Six checklists; one app.
\end{itemize}
\end{recap}

\section*{Workshop}

\begin{enumerate}[leftmargin=*]
  \item Run the checklist on your current project.
  \item Note what's missing; fix or document.
  \item Ship.
\end{enumerate}
